---
title: "Selecting OTUs"
author: "Cynthia Rodriguez"
output: html_document
---

-----

### Import data:
```{r}
data1 = read.csv("Input_files/Baxter_deseq2_result_table.csv", header=TRUE, sep=",", skipNul=TRUE)

data2 = read.csv("Input_files/Healey_deseq2_result_table.csv", header=TRUE, sep=",", skipNul=TRUE)

data3 = read.csv("Input_files/Hooda_deseq2_result_table.csv", header=TRUE, sep=",", skipNul=TRUE)

data4 = read.csv("Input_files/Liu_deseq2_result_table.csv", header=TRUE, sep=",", skipNul=TRUE)

data5 = read.csv("Input_files/Morales_deseq2_result_table.csv", header=TRUE, sep=",", skipNul=TRUE)

data6 = read.csv("Input_files/Tap_deseq2_result_table.csv", header=TRUE, sep=",", skipNul=TRUE)

data7 = read.csv("Input_files/Vandeputte_deseq2_result_table.csv", header=TRUE, sep=",", skipNul=TRUE)

data8 = read.csv("Input_files/Venkataraman_deseq2_result_table.csv", header=TRUE, sep=",", skipNul=TRUE)

```

### Extract OTU column as a vector:
```{r}
data1vector <- c(data1$OTU)
 
data2vector <- c(data2$OTU)
 
data3vector <- c(data3$OTU)
 
data4vector <- c(data4$OTU)
 
data5vector <- c(data5$OTU)
 
data6vector <- c(data6$OTU)
 
data7vector <- c(data7$OTU)
 
data8vector <- c(data8$OTU)

```


### Create one vector with OTUs from all studies:
```{r}
mergedvector <- c(data1vector, data2vector, data3vector, data4vector, data5vector, data6vector, data7vector, data8vector)

#Save merged vector:
write.csv(mergedvector, file = "Output_files/1.all.OTUs.csv")
```

### Find all the OTUs:
```{r}
#How many characters repeat:
length(which(table(mergedvector)>1))

#Which characters repeat and their position in the vector:
all_OTUs <-which(table(mergedvector)>1)

#Save characters that repeat
write.csv(all_OTUs, file = "Output_files/2.present.all_OTUs.csv")
```

### Find OTUs that repeat in all studies:
```{r}
#How many characters repeat:
length(which(table(mergedvector)>7))

#Which characters repeat and their position in the vector:
all_repeated_OTUs <-which(table(mergedvector)>7)

#Save characters that repeat
write.csv(all_repeated_OTUs, file = "Output_files/3.repeated_OTUs_all.csv")
```


### Find OTUs that repeat more 5 times or more:
```{r}
#How many characters repeat:
length(which(table(mergedvector)>4))

#Which characters repeat and their position in the vector:
repeated_OTUs_5 <-which(table(mergedvector)>4)

#Save characters that repeat
write.csv(repeated_OTUs_5, file = "Output_files/4.repeated_OTUs_>5.csv")
```

### Find OTUs that repeat 3 times or more:
```{r}
#How many characters repeat:
length(which(table(mergedvector)>2))

#Which characters repeat and their position in the vector:
repeated_OTUs_5 <-which(table(mergedvector)>2)

#Save characters that repeat
write.csv(repeated_OTUs_5, file = "Output_files/4.repeated_OTUs_>3.csv")
```


Results:
272 OTUs repeat 3 times or more.

### Extract rows with repeated_OTUs identified from raw data:
```{r}
library("dplyr")

#Import data with the repeated OTUs: change header to say OTU before importing
repeated_OTUs1 = read.csv("Output_files/4.repeated_OTUs_>3.csv", header=TRUE, sep=",", skipNul=TRUE)
head(repeated_OTUs1)

repeated_OTUs_vector <- c(repeated_OTUs1$OTU)

#See vector:
head(repeated_OTUs_vector)

#Baxter:
#Extract rows based on vector imported:
data1_repeated_5 <-data1[data1$OTU %in% repeated_OTUs_vector,]

#Save data in file::
write.table(data1_repeated_5, file = "Output_files/5.Baxter_repeated_OTUs_>3.csv", sep = ",", row.names = FALSE, col.names = TRUE)


#Healey:
#Extract rows based on vector imported:
data2_repeated_5 <-data2[data2$OTU %in% repeated_OTUs_vector,]

#Save data in file::
write.table(data2_repeated_5, file = "Output_files/5.Healey_repeated_OTUs_>3.csv", sep = ",", row.names = FALSE, col.names = TRUE)


#Hooda:
#Extract rows based on vector imported:
data3_repeated_5 <-data3[data3$OTU %in% repeated_OTUs_vector,]

#Save data in file::
write.table(data3_repeated_5, file = "Output_files/5.Hooda_repeated_OTUs_>3.csv", sep = ",", row.names = FALSE, col.names = TRUE)


#Liu:
#Extract rows based on vector imported:
data4_repeated_5 <-data4[data4$OTU %in% repeated_OTUs_vector,]

#Save data in file::
write.table(data4_repeated_5, file = "Output_files/5.Liu_repeated_OTUs_>3.csv", sep = ",", row.names = FALSE, col.names = TRUE)


#Morales:
#Extract rows based on vector imported:
data5_repeated_5 <-data5[data5$OTU %in% repeated_OTUs_vector,]

#Save data in file::
write.table(data5_repeated_5, file = "Output_files/5.Morales_repeated_OTUs_>3.csv", sep = ",", row.names = FALSE, col.names = TRUE)


#Tap:
#Extract rows based on vector imported:
data6_repeated_5 <-data6[data6$OTU %in% repeated_OTUs_vector,]

#Save data in file::
write.table(data6_repeated_5, file = "Output_files/5.Tap_repeated_OTUs_>3.csv", sep = ",", row.names = FALSE, col.names = TRUE)


#Vandeputte:
#Extract rows based on vector imported:
data7_repeated_5 <-data7[data7$OTU %in% repeated_OTUs_vector,]

#Save data in file::
write.table(data7_repeated_5, file = "Output_files/5.Vandeputte_repeated_OTUs_>3.csv", sep = ",", row.names = FALSE, col.names = TRUE)


#Venkataraman:
#Extract rows based on vector imported:
data8_repeated_5 <-data8[data8$OTU %in% repeated_OTUs_vector,]

#Save data in file::
write.table(data8_repeated_5, file = "Output_files/5.Venkataraman_repeated_OTUs_>3.csv", sep = ",", row.names = FALSE, col.names = TRUE)
```

### Merge dataframes that have common OTUs (present in at least 5 studies):
```{r}
Merged <- rbind(data1_repeated_5, data2_repeated_5, data3_repeated_5, data4_repeated_5, data5_repeated_5, data6_repeated_5, data7_repeated_5, data8_repeated_5)

head(Merged)

#Save merged data:
write.table(Merged, file = "Output_files/6.Merged_repeated_OTUs_>3.csv", sep = ",", row.names = FALSE, col.names = TRUE)

#Outside of R sort the data by OTU

```


### Average across OTUs from merged data (present in at least 5 studies):
Reference:https://www.geeksforgeeks.org/how-to-calculate-the-mean-by-group-in-r-dataframe/
```{r}
#Import merged data with the repeated OTUs:
Merged1 = read.csv("Output_files/6.Merged_repeated_OTUs_>3_sorted.csv", header=TRUE, sep=",", skipNul=TRUE)
head(Merged1)

#Average across repeated OTUs:
average_OTUs <- aggregate(x= Merged1$log2FoldChange,     
                 by = list(Merged1$OTU),   # Specify group indicator    
                  FUN = mean)   # Specify function (i.e. mean)

#Save merged data with averages:
write.table(average_OTUs, file = "Output_files/7.Merged_repeated_OTUs_averages_>3.csv", sep = ",", row.names = FALSE, col.names = TRUE)
```

### Add taxonomic assignment to the log2FoldChange averages: 
```{r}
#Created a taxonomy file in excel by only having OTU column and taxonomy columns - import it:
Taxonomy = read.csv("Output_files/7.Taxonomy_OTUs_>3_sorted.csv", header=TRUE, sep=",", skipNul=TRUE)

#Combine average_OTUs with taxonomy:
average_OTUs_taxonomy <- cbind(average_OTUs, Taxonomy)

#Rename columns in for OTU and log2FoldChange:
as.factor(average_OTUs_taxonomy$Group.1)
as.factor(average_OTUs_taxonomy$x)

names(average_OTUs_taxonomy)[names(average_OTUs_taxonomy) == "Group.1"] <- "OTU1"
names(average_OTUs_taxonomy)[names(average_OTUs_taxonomy) == "x"] <- "log2FoldChange"

#Save merged data with averages and taxonomy:
write.table(average_OTUs_taxonomy, file = "Output_files/8.Merged_repeated_OTUs_averages_taxonomy_>3.csv", sep = ",", row.names = FALSE, col.names = TRUE)

```


