---
title: "Betadisper Deehan_tapioca"
---

```{=html}
<style>

body, td {
   font-family: "Tw Cen MT";
   font-size: 16px;
}

</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

### Packages needed:
```{r}
require(vegan)
require(ggplot2)
require(data.table)
require(anchors)
require(EcolUtils)
require(car)
require(tidyverse)
require(dplyr)
require(rstatix)
library("extrafont")
```

### Set working directory:
```{r}
setwd("/Volumes/Seagate_Cedars/Martiny_Lab/Meta-analysis/Data_analysis/Meta_analysis/Betadisp/Deehan_2020_tapioca")
```

## __________ Documents needed: __________##

* 1. *beta_matrix* - **9.tapioca_median_bray_curtis_sqrt_q1_forward_rarefied_to_7157.txt** 

* 2. *metadata* - **1.Deehan_metadata_tapioca.tsv** 

* 3. *beta_metrics_merged_metadata* - **8.NMDS.coordinates.metadata.txt** 

## __________ Read in your data _________ ##


#### Read in table with all alpha diversity data:
```{r}
beta_metrics_merged_metadata <- as.data.frame(fread("Input/8.NMDS.coordinates.metadata.txt", header = TRUE))
str(beta_metrics_merged_metadata)

beta_matrix <- as.data.frame(read.table("Input/9.tapioca_median_bray_curtis_sqrt_q1_forward_rarefied_to_7157.txt", header = TRUE, row.names=1))


metadata <- as.data.frame(fread("Input/1.Deehan_metadata_tapioca.tsv", header = TRUE))
```


## ___________ Beta-Diversity BETADISPER Analysis ________ ##

### Test for variation dispersion of the data:
```{r}

#Run betadisper test:
betadisper.test <- betadisper(as.dist(beta_matrix), metadata$fiber_amount2, type = "centroid", bias.adjust = FALSE,
       sqrt.dist = FALSE, add = FALSE)

#Check results:
set.seed(2022)
bd.result <- permutest(betadisper.test)
bd.result
```

Result: No significant variation between the before and after samples.

##### Save PERMDISP results:
```{r}
  write.table(x = bd.result$tab, file = "Output/1.Betadisper_1.txt", sep = "\t", row.names = TRUE, col.names = NA) 
```

### Confirm PERMANOVA result from before:
```{r}
set.seed(2022) #reproducible results
  PERMANOVA1 <- adonis2(as.dist(beta_matrix)~`subject_id`*`fiber_amount2`, data=metadata, method = "bray", by="terms",permutations = 999)
  PERMANOVA1
```

```{r}
set.seed(2022) #reproducible results
PERMANOVA3 <-  adonis2(as.dist(beta_matrix)~`subject_id`+`fiber_amount2`, data=metadata, method = "bray", by="terms",permutations = 999)
PERMANOVA3
```


### NMDS plot:
```{r}
NMDS.timepoint <- ggplot(data = beta_metrics_merged_metadata) +
  scale_fill_manual(values=c("#900C3F", "#FFC300")) +
  aes(x = MDS1, y = MDS2, fill = as.factor(`fiber_amount2`)) + 
  geom_point(pch = 21, aes(fill = as.factor(`fiber_amount2`)), size = 5, alpha = 0.7) +
labs(title = "NMDS Deehan_tapioca", x= "MDS1", y= "MDS2", fill="Timepoint") +
  stat_ellipse(size=1,aes(color=`fiber_amount2`)) +
  scale_color_manual(values=c("#900C3F", "#FFC300")) +
  theme_bw (base_family = "Tw Cen MT", base_size = 12) +
  guides (fill=guide_legend(nrow=9))

NMDS.timepoint
```

#### Save figure:
```{r}
ggsave("Figures/NMDS.Deehan_tapiocae.png", width = 7, height = 5)
```


##### To see session information:
```{r}
sessionInfo()
```
